---
title: Template for Oxford University Press papers
author:
  - name: Alpha Beta
    email: alpha@university.ac.uk
    corresponding_author: TRUE # Automatic footnote using corresponding_author
    affiliation:
      - ABC
      - DEF
    biography: |
      A researcher.
      #    biography_pic: "/PATH/TO/image.png"
      #    biography_pic_width: "1in"
  - name: Beta Gamma
    email: beta@university.ac.uk
    affiliation: DEF
    footnote: 1 # Footnote using reference
  - name: Zeta Epsilon
    email: zeta@institution.ac.uk
    affiliation: XYZ
    footnote:
        - 1
        - 2
address:
  - code: ABC
    department: Department of Biology
    organization: University
  - code: DEF
    address: University
  - code: XYZ
    address: Institution
footnote:
  - code: 1
    text: "Equal contribution"
  - code: 2
    text: "Current email address: [alpha@university.ac.uk](mailto:alpha@university.ac.uk)"
abstract:
  - "**Objective:**  To Investigate, and identify, the factors that may influence patient survival of coronavirus disease 19."
  - "More"
  - "**Motivation:** You can also have some paragraphs start with bold face."
keywords:
  - mathematics
  - survival
  - prognostics
competing_interests: |
  There are no competing interest.
author_contributions: |
  Equal contributions
acknowledgements: |
  This is an acknowledgement.

  It consists of two paragraphs.

bibliography: manuscript/mybibfile.bib # Includes refs in OUP example template
# When using `citation_package = "default"`, a CSL file can be used
# csl: https://www.zotero.org/styles/oxford-university-press-note
output:
  rticles::oup_article:
    oup_version: 1 # 1 = 2020 CTAN OUP CLS package; 0 = 2009 OUP CLS package
    keep_tex: FALSE
    #journal: "Journal Name"
    document_style: "contemporary" # Can be contemporary, modern, traditional
    papersize: "large" # Can be large, medium, small
    #citation_package: "default" # Uncomment when using a CSL; default "natbib"
    namedate: TRUE # Set FALSE to use numeric refs; Default FALSE
    #number_sections: FALSE # Uncomment to not number sections; default TRUE
    #number_lines: TRUE # Use `lineno` package - Default FALSE
    #number_lines_options: ["mathlines","switch"]  # Options for latex lineno package.
    #onecolumn: TRUE # Uncomment for one column format; default FALSE
    extra_dependencies:
      - booktabs # to use with knitr::kable() example below
      
# Example of pandoc's variable
# urlcolor: orange
# linkcolor: green
# citecolor: red
header-includes:
  - \theoremstyle{thmstyleone}
  - \newtheorem{theorem}{Theorem}
  - \newtheorem{proposition}[theorem]{Proposition}
  - \theoremstyle{thmstyletwo}
  - \newtheorem{example}{Example}
  - \newtheorem{remark}{Remark}
  - \theoremstyle{thmstylethree}
  - \newtheorem{definition}{Definition}
---

<!--- Global Settings
    th -> top or here
--->
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.pos = 'th',
  out.width = '100%', dpi = 300, fig.env="figure")

options(knitr.table.format = "latex") # For kable tables to write LaTeX table directly
```

<!--- imports --->
```{r include = FALSE}
library(xtable)
library(dplyr)
library(magrittr)
```

<!--- external functios --->
```{r}
sys.source(file = 'R/functions/StudyData.R', envir = knitr::knit_global())
```


# Introduction

This template is based on the generic OUP authoring template available on CTAN under [oup-authoring-template](https://www.ctan.org/pkg/oup-authoring-template). The CTAN template includes LaTeX documentation and a sample LaTeX document that provide far more details regarding the full functionality of the format. Here, only the basic functioning of the Rmarkdown adaptation of the format is demonstrated.

## A subsection

A numbered list:

1) First point
2) Second point
    - Subpoint
    
A bullet list:

* First point
* Second point

## Notes

* Extra white space in document will tend to disappear as text is filled in.
* Code blocks tend to generate lots of empty white space when `echo=TRUE` for some reason.



# Methods

## Data Source
The data is a derivative of the ... It is a cohort study wherein ...

## Participants

* Study setting
* Eligibility criteria

## Outcome

* Died

## Predictors






# Literature citations

By default, citations are handled by `natbib` using a numeric citation format. To use name-date citations, sets `namedate: TRUE` in the YAML header.

Here are two sample references: 

* **author (year) example:** @horvath2018dna showed some really cool things. Only seems to work properly if `namedate: TRUE`. 
* **(author year) example:** This is a well known result [@ji20123d]. 

The bibliography will appear at the end of the document.

Though not normally available in the OUP LaTeX format, [CSL style files](https://www.zotero.org/styles) can also be used with the Rmarkdown adaptation by setting in the YAML header `citation_package: "default"` and defining the `csl` element to be the path towards the style file.

# Equations

An equation without a label for cross-referencing:

$$
E=mc^2
$$

An inline equation: $y=ax+b$

An equation with a label for cross-referencing:

\begin{equation}\label{eq:eq1}
\int^{r_2}_0 F(r,\varphi){\rm d}r\,{\rm d}\varphi = 1
\end{equation}

This equation can be referenced as follows: Eq. \ref{eq:eq1}

# Inserting R figures

The code below creates a figure.

```{r fig1, fig.cap="This is the first figure."}
plot(1:10,main="Some data",xlab="Distance (cm)",
     ylab="Time (hours)")
```

You can reference this figure as follows: Fig. \ref{fig:fig1}.

## Figures spanning two-columns

Figures can span two columns be setting `fig.env="figure*"`.

```{r fig2, fig.cap="This is a wide figure.", fig.env="figure*"}
plot(1:5,pch=19,main="Some wide data",
     xlab="Distance (cm)",ylab="Time (hours)")
```

Reference to second figure: Fig. \ref{fig:fig2}

# Tables

## Generate a table using `xtable`

```{r xtabletable, results="asis", echo=TRUE}
df <- data.frame(ID=1:3,code=letters[1:3])

# Creates tables that follow OUP guidelines 
# using xtable

print(xtable::xtable(df,caption="This is a xtable table.", label="tab:tab1"), comment=FALSE,caption.placement="top")
```

You can reference this table as follows: Table \ref{tab:tab1}.

## Generate a table using `kable`

```{r kabletable, echo=TRUE}
df <- data.frame(ID=1:3,code=letters[1:3])

# kable can alse be used for creating tables
knitr::kable(df,caption="This is a kable table.",
             booktabs=TRUE,label="tab2")
```

You can reference this table as follows: Table \ref{tab:tab2}.

## Table spanning two columns

Tables can span two columns be setting `table.envir = "table*"` in `knitr::kable`.

```{r widetable, echo=TRUE}
df <- data.frame(ID=1:3,code1=letters[1:3],
                code2=letters[4:6],
                code3=letters[7:9],
                code4=letters[10:12],
                code5=letters[13:15])

# kable can alse be used for creating tables
knitr::kable(df,caption="This is a wide kable table.",
             #format="latex",
             table.envir="table*",
             booktabs=TRUE,label="tab3")
```




# Cross-referencing sections

You can cross-reference sections and subsections as follows: Section \ref{literature-citations} and Section \ref{a-subsection}.

***Note:*** the last section in the document will be used as the section title for the bibliography.

For more portable and flexible referencing of sections, equations, figures and tables, use [`bookdown::pdf_document2`](https://github.com/rstudio/bookdown) with YAML header option `base_format: rticles::oup_article`.

# Appendices {-}

::: {.appendices latex=true}

# Section title of first appendix

blabla

## Subsection title of first appendix

and so on....
:::

# References
